# Reproducible Research  Assessment 1

## Loading and preprocessing the data
```{r loaddata}
setwd("C:/WorkSpace/DataScience/5.ReproducibleResearch/Project/Assignment1")
activity <- read.csv("Data/activity.csv")
activity$date <- as.Date(activity$date)
```
## What is mean total number of steps taken per day?
```{r}
# Load reshape2 library to get melt & dcast functions
library(reshape2)

# Melt data frame to prep for casting by date.
# By setting the id variable to date and the measure variable to steps, 
# we get a table with multiple values for steps taken within each day
actMeltDate <- melt(activity, id.vars="date", measure.vars="steps", na.rm=FALSE)

# Cast data frame to see steps per day -- this sums the steps by date 
# to give us a table of 3 columns x 61 rows
actCastDate <- dcast(actMeltDate, date ~ variable, sum)

# Plot histogram with frequency of steps by day and add a red line 
# showing the mean value
plot(actCastDate$date, actCastDate$steps, type = "h", 
     main="Histogram of Daily Steps", xlab="Date", 
     ylab="Steps per Day", col="blue", lwd=8)
abline(h=mean(actCastDate$steps, na.rm=TRUE), col="red", lwd=2)

# Calculate mean and median of daily steps
paste("Mean Steps per Day =", mean(actCastDate$steps, na.rm=TRUE))
paste("Median Steps per Day =", median(actCastDate$steps, na.rm=TRUE))
```

## What is the average daily activity pattern?
```{r}
# Re-melt data frame to prep for casting by interval, 
# including removing NA values 
# so we can take the mean a little later
actMeltInt <- melt(activity, id.vars="interval", 
                   measure.vars="steps", na.rm=TRUE)

# Cast data frame to see mean steps per interval
actCastInt <- dcast(actMeltInt, interval ~ variable, mean)

# Plot line chart with average frequency of steps 
# by interval and add line with mean
plot(actCastInt$interval, actCastInt$steps, type ="l", 
     main="Frequency of Steps Taken at Each Interval", 
     xlab="Interval ID", ylab="Steps", col="orange", lwd=2)
abline(h=mean(actCastInt$steps, na.rm=TRUE), col="red", lwd=2)

# The plot shows the peak at somewhere in the 800-900 
# interval range so let's find out exactly which interval 
# has the max value and what that maximum value is
# Output interval that has max value along with the max value
paste("Interval with max value =", 
      actCastInt$interval[which(actCastInt$steps == max(actCastInt$steps))])
paste("Maximum interval mean steps =", max(actCastInt$steps))

```

## Imputing missing values

There are a number of days/intervals where there are missing 
values (coded as NA). The presence of missing days may introduce 
bias into some calculations or summaries of the data.

```{r MissingValueNumber}
# Calculate number of rows in activity data set with NA rows
sum(is.na(activity$steps))
```

All of the missing values are filled in with mean 
value of its 5-minute interval.

```{r}
# Data frame with mean steps per interval - just renaming to be more descriptive
stepsPerInt <- actCastInt

# Create data frame that we will remove NAs from
actNoNA <- activity

# Merge activity data set with stepsPerInt data set
actMerge = merge(actNoNA, stepsPerInt, 
                 by="interval", suffixes=c(".act", ".spi"))

# Get list of indexes where steps value = NA
naIndex = which(is.na(actNoNA$steps))

# Replace NA values with value from steps.spi
actNoNA[naIndex,"steps"] = actMerge[naIndex,"steps.spi"]
```
Histogram of the total number of steps taken 
each day after missing values are imputed
```{r}
# Melt new data frame to prep for casting by date
actMeltDateNoNA <- melt(actNoNA, id.vars="date", 
                        measure.vars="steps", na.rm=FALSE)

# Cast data frame to see steps per day
actCastDateNoNA <- dcast(actMeltDateNoNA, date ~ variable, sum)

# Plot histogram with frequency of steps by day
plot(actCastDateNoNA$date, actCastDateNoNA$steps, 
     type="h", main="Histogram of Daily Steps (Imputted NA Values)", 
     xlab="Date", ylab="Steps", col="gray", lwd=8)
abline(h=mean(actCastDateNoNA$steps), col="red", lwd=2)

# Calculate mean and median of daily steps
paste("Mean daily steps =", mean(actCastDateNoNA$steps, na.rm=TRUE))

paste("Median daily steps =", median(actCastDateNoNA$steps, na.rm=TRUE))
```

Mean and median values are higher after imputing missing data. The reason is
that in the original data, there are some days with `steps` values `NA` for 
any `interval`. The total number of steps taken in such days are set to 0s by
default. However, after replacing missing `steps` values with the mean `steps`
of associated `interval` value, these 0 values are removed from the histogram
of total number of steps taken each day.

## Are there differences in activity patterns between weekdays and weekends?
First, let's find the day of the week for each measurement in the dataset. In
this part, we use the dataset with the filled-in values.

```{r}
# For loop to create new column called "dayOfWeek" and 
# insert whether each date corresponds to a weekday or weekend
for (i in 1:nrow(actNoNA)) {
    if (weekdays(actNoNA$date[i]) == "Saturday" | 
        weekdays(actNoNA$date[i]) == "Sunday") {
        actNoNA$dayOfWeek[i] = "weekend"
    } else {
        actNoNA$dayOfWeek[i] = "weekday"
    }
}

# To create a plot, we must first subset the data
actWeekday <- subset(actNoNA, dayOfWeek=="weekday")
actWeekend <- subset(actNoNA, dayOfWeek=="weekend")

# Next, we need to process the data for our needs
actMeltWeekday <- melt(actWeekday, id.vars="interval", measure.vars="steps")
actMeltWeekend <- melt(actWeekend, id.vars="interval", measure.vars="steps")
actCastWeekday <- dcast(actMeltWeekday, interval ~ variable, mean)
actCastWeekend <- dcast(actMeltWeekend, interval ~ variable, mean)

# Load plot packages necessary to continue
library(ggplot2)
library(gridExtra)
```

Now, let's make a panel plot containing plots of average 
number of steps taken
on weekdays and weekends.
```{r}
# Set plot area to two rows and one column, and then plot 
# charts with mean line in each
plot1 <- qplot(actCastWeekday$interval, actCastWeekday$steps, 
               geom="line", data=actCastWeekday, main="Steps by Interval - Weekday", xlab="Interval ID", ylab="Number of Steps")
plot2 <- qplot(actCastWeekend$interval, actCastWeekend$steps, 
               geom="line", data=actCastWeekend, main="Steps by Interval - Weekend", xlab="Interval ID", ylab="Number of Steps")
grid.arrange(plot1, plot2, nrow=2)
```

